---
title: "scRNA"
author: "aidan borkan"
date: "2026-02-21"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
# installation
#install.packages("Matrix")
#BiocManager::install("GEOquery")
# ensure over 1.1.7
#install.packages("rlang", type = "binary")
#packageVersion("rlang")
reticulate::py_install("umap-learn", pip = TRUE)

# Comprehensive list
library(GEOquery)
library(dplyr)
library(Seurat)
library(patchwork)
library(data.table)
library(Matrix)
library(enrichplot)
library(stringr)
library(ggplot2)
# Annotations
library(SingleR)
library(celldex)
library(SingleCellExperiment)

# pathway mapping
#install.packages("clusterProfiler")
library(clusterProfiler)
library(org.Hs.eg.db)

gse <- getGEO("GSE75140", GSEMatrix = TRUE)
expr <- exprs(gse[[1]])
```

```{r inspect , message=TRUE, warning=FALSE}
# gse is your GEO object (from getGEO)
eset <- gse[[1]]

#expr <- exprs(eset)      
# expression matrix
pheno <- pData(eset)     
# sample metadata
fdata <- fData(eset)     
# feature metadata (genes)
dim(expr)     
# genes x cells 
dim(pheno)
head(colnames(pheno))
```
# often on GEO we need the supplementary data for expr
```{r suppl, message=TRUE, warning=FALSE}
path <- "C:/Users/suzan/Downloads/GSE75140_hOrg.fetal.master.data.frame.txt.gz"

# -----------------------------
# 1) Load
# -----------------------------
df <- fread(
  path,
  sep = "\t",
  header = TRUE,
  data.table = FALSE,
  check.names = FALSE
)
cat("Loaded df:", nrow(df), "rows x", ncol(df), "cols\n")

# -----------------------------
# 2) Set cell IDs (rownames)
# -----------------------------
id_candidates <- c("cell_id","cell","Cell","barcode","Barcode","cell_name","cellname",
                   "cellID","CellID","cellid","X","...1","V1")
id_col <- intersect(id_candidates, colnames(df))
id_col <- if (length(id_col) >= 1) id_col[1] else colnames(df)[1]

x <- df[[id_col]]
looks_like_id <- (is.character(x) || is.factor(x)) &&
  length(unique(x)) == nrow(df) &&
  !any(is.na(x))

if (looks_like_id) {
  cat("Using cell id column:", id_col, "\n")
  rownames(df) <- as.character(x)
  df[[id_col]] <- NULL
} else {
  cat("No usable cell-id column found; creating Cell_1..Cell_n\n")
  rownames(df) <- paste0("Cell_", seq_len(nrow(df)))
}
stopifnot(length(rownames(df)) == nrow(df))

# -----------------------------
# 3) Decide expression vs metadata (CORRECT FOR SPARSE scRNA)
#    Key idea: genes are numeric columns. Many genes are low-cardinality because zeros dominate.
#    So DO NOT use "unique > 50" for genes.
# -----------------------------
sample_n <- min(50, nrow(df))
idx <- seq_len(sample_n)

is_mostly_numeric <- vapply(df, function(col) {
  xn <- suppressWarnings(as.numeric(col[idx]))
  mean(!is.na(xn)) >= 0.95
}, logical(1))

# Metadata whitelist by NAME (keep this tight)
meta_name_pat <- paste(
  c(
    "^orig\\.ident$",
    "^sample(_id)?$",
    "^donor$",
    "^patient$",
    "^subject$",
    "^batch$",
    "^chip$",
    "^lane$",
    "^run$",
    "^replicate$",
    "^rep$",
    "^time(point)?$",
    "^day$",
    "^week$",
    "^stage$",
    "^tissue$",
    "^region$",
    "^condition$",
    "^group$",
    "^treatment$",
    "^sex$",
    "^age$",
    "GSE",
    "GSM",
    "geo",
    "accession"
  ),
  collapse = "|"
)

is_meta_by_name <- grepl(meta_name_pat, colnames(df), ignore.case = TRUE)

# Expression columns = numeric-ish and NOT named like metadata
expr_cols <- names(df)[is_mostly_numeric & !is_meta_by_name]
meta_cols <- setdiff(names(df), expr_cols)

cat("Expression cols:", length(expr_cols), "\n")
cat("Metadata cols:", length(meta_cols), "\n")

# sanity: metadata should not be mostly gene symbols
bad_meta_like_genes <- meta_cols[grepl("^[A-Z0-9]+(-[A-Z0-9]+)?$", meta_cols)]
cat("Metadata cols that LOOK like gene symbols (should be small):", length(bad_meta_like_genes), "\n")
if (length(bad_meta_like_genes) > 0) print(head(bad_meta_like_genes, 20))

meta_df <- df[, meta_cols, drop = FALSE]
expr_df <- df[, expr_cols, drop = FALSE]

cat("First metadata columns:\n")
print(head(meta_cols, 50))

# -----------------------------
# 4) Convert expression -> numeric matrix (genes x cells)
# -----------------------------
expr_mat_cells_by_genes <- as.matrix(expr_df)
expr_mat_cells_by_genes <- apply(expr_mat_cells_by_genes, 2, function(v) suppressWarnings(as.numeric(v)))
rownames(expr_mat_cells_by_genes) <- rownames(expr_df)

na_frac <- mean(is.na(expr_mat_cells_by_genes))
cat("NA fraction after numeric conversion:", signif(na_frac, 4), "\n")
stopifnot(na_frac == 0)

mat <- t(expr_mat_cells_by_genes)  # genes x cells
cat("Expression range:", range(mat), "\n")

# -----------------------------
# 5) Build Seurat obj
# -----------------------------
zero_counts <- Matrix(0, nrow = nrow(mat), ncol = ncol(mat), sparse = TRUE)
rownames(zero_counts) <- rownames(mat)
colnames(zero_counts) <- colnames(mat)

obj <- CreateSeuratObject(counts = zero_counts, meta.data = meta_df)
obj <- SetAssayData(obj, assay = "RNA", layer = "data", new.data = mat)

# -----------------------------
# 6) Inspect metadata candidates for sample_id / condition
# -----------------------------
md <- obj@meta.data
cat("Metadata cols in Seurat:", ncol(md), "\n")
print(colnames(md))

uniq_md <- sort(sapply(md, function(col) length(unique(col))))
cat("\nMetadata columns sorted by #unique values:\n")
print(head(uniq_md, 30))

# Show frequency tables for columns with smallish number of levels
cand <- names(uniq_md[uniq_md > 1 & uniq_md <= 30])
if (length(cand) == 0) {
  cat("\nNo categorical metadata columns found (uniq between 2 and 30).\n")
} else {
  cat("\nCandidate categorical metadata columns (2..30 uniques):\n")
  print(cand)
  for (nm in head(cand, 10)) {
    cat("\n---", nm, "--- uniques:", uniq_md[[nm]], "\n")
    print(head(table(md[[nm]]), 30))
  }
}
```

# differentiate samples
```{r message=FALSE, warning=FALSE}
# index of species

obj_sub <- subset(obj, subset = species %in% c("hOrg","human"))
Idents(obj_sub) <- "species"
table(Idents(obj_sub))
```



# inspect distribution to find percentiles that can guide cutoffs
```{r message=FALSE, warning=FALSE}

# Add mitochondrial % to metadata
# 1) percent.mt (optional)
# Only makes sense if gene names include MT- genes

# ---- Use data layer for QC proxies ----
data_mat <- GetAssayData(obj, assay = "RNA", layer = "data")

# genes detected proxy (logexpr > 0)
obj$genes_detected <- Matrix::colSums(data_mat > 0)

# library size proxy (sum of logexpr)
obj$sum_logexpr <- Matrix::colSums(data_mat)

# MT signal (robust)
mt_genes <- grep("^MT-", rownames(obj), value = TRUE, ignore.case = TRUE)
if (length(mt_genes) > 0) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-", assay = "RNA")
  obj$mt_score <- Matrix::colMeans(data_mat[mt_genes, , drop = FALSE])
} else {
  obj$percent.mt <- NA_real_
  obj$mt_score <- NA_real_
  message("No MT- genes detected; percent.mt and mt_score set to NA.")
}

# ---- HVGs (pick ONE approach) ----
# Option 1 (simple): Seurat HVGs
#  DefaultAssay(obj) <- "RNA"
# obj <- FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000)

# Option 2 (if you want purely variance on data layer, sparse-safe):
data_mat <- GetAssayData(obj, assay="RNA", layer="data")

gene_means  <- Matrix::rowMeans(data_mat)
gene_meansq <- Matrix::rowMeans(data_mat^2)
gene_var    <- gene_meansq - gene_means^2

VariableFeatures(obj) <- names(sort(gene_var, decreasing=TRUE))[1:2000]

# ---- downstream ----
obj <- ScaleData(obj, features = VariableFeatures(obj))
obj <- RunPCA(obj, features = VariableFeatures(obj))
obj <- RunUMAP(obj, dims = 1:20, umap.method = "uwot", metric = "cosine")
obj <- FindNeighbors(obj, dims = 1:20)
obj <- FindClusters(obj)

# quick QC summaries
quantile(obj$genes_detected, probs = c(0.01,0.05,0.1,0.5,0.9,0.95,0.99), na.rm = TRUE)
quantile(obj$sum_logexpr,  probs = c(0.01,0.05,0.1,0.5,0.9,0.95,0.99), na.rm = TRUE)
if (any(!is.na(obj$mt_score))) {
  quantile(obj$mt_score, probs = c(0.5,0.9,0.95,0.99), na.rm = TRUE)
}
```


# Proxy for standard QC
```{r}

# proxy
plot_feats <- c("genes_detected", "sum_logexpr")
VlnPlot(obj, features = plot_feats, ncol = length(plot_feats))

# note, geo data matrix does not have mt 
# higher genes_detected means higher transcriptomic complexity
```
# example of how to test if a cluster (3) is biologically meaningful: cluster and plot UMAP and see if 3 is an edge island or randomly scattered fprminmg a diffuse cloud
```{r}
FeaturePlot(obj, features = c("genes_detected"))
DimPlot(obj, label = TRUE)
```

# follow : https://satijalab.org/seurat/articles/pbmc3k_tutorial
```{r}


# ------------------------------------------------------------
# 3) Scale on HVGs only
# ------------------------------------------------------------

obj <- ScaleData(obj, features = VariableFeatures(obj), vars.to.regress = "sum_logexpr", verbose = FALSE)

# ------------------------------------------------------------
# 4) PCA
# ------------------------------------------------------------
obj <- RunPCA(obj, features = VariableFeatures(obj), verbose = FALSE)
ElbowPlot(obj, ndims = 50)

```
# visually flattens around PC12 - PC15
```{r}
# pick dims after elbow
nd <- 15

# ------------------------------------------------------------
# 5) Neighbors + Clusters
# ------------------------------------------------------------
obj <- FindNeighbors(obj, dims = 1:nd, verbose = FALSE)
obj <- FindClusters(obj, resolution = 0.4, verbose = FALSE)

# ------------------------------------------------------------
# 6) UMAP
# ------------------------------------------------------------
obj <- RunUMAP(obj, dims = 1:nd, umap.method = "uwot", metric = "cosine", verbose = FALSE)
DimPlot(obj, reduction = "umap", label = TRUE)
```


```{r}
VizDimLoadings(obj, dims = 1:2, reduction = "pca")

```

```{r}
DimPlot(obj, reduction = "pca") + NoLegend()
```

```{r}
DimHeatmap(obj, dims = 1, cells = 500, balanced = TRUE)

```
# use dimensions PC1 - PC15
```{r}
obj <- RunTSNE(obj, dims = 1:15)
DimPlot(obj, reduction = "tsne", label = TRUE)
```
#let's find markers
```{r}
markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

# more strict across all rows
markers2 <- markers %>%
  group_by(cluster) %>%
  mutate(p_adj_clusterBH = p.adjust(p_val, method = "BH")) %>%  # explicit per-cluster BH
  ungroup() %>%
  mutate(p_adj_globalBH = p.adjust(p_val, method = "BH"))       # global BH across all tests

# sanity: confirm adjusted p-values exist
stopifnot("p_val_adj" %in% colnames(markers))
all.equal(markers2$p_val_adj, markers2$p_adj_clusterBH)
library(dplyr)

# how many markers
markers2 %>%
  summarize(
    n_total = n(),
    n_sig_seurat = sum(p_val_adj < 0.05, na.rm = TRUE),
    n_sig_clusterBH = sum(p_adj_clusterBH < 0.05, na.rm = TRUE),
    n_sig_globalBH = sum(p_adj_globalBH < 0.05, na.rm = TRUE)
  )

# per cluster
markers2 %>%
  group_by(cluster) %>%
  summarize(
    n_tests = n(),
    sig_seurat = sum(p_val_adj < 0.05, na.rm = TRUE),
    sig_globalBH = sum(p_adj_globalBH < 0.05, na.rm = TRUE)
  ) %>%
  arrange(desc(sig_seurat))

top_seurat <- markers2 %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  ungroup()

top_global <- markers2 %>%
  filter(p_adj_globalBH < 0.05) %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  ungroup()

key_seurat <- paste(top_seurat$cluster, top_seurat$gene, sep = "::")
key_global <- paste(top_global$cluster, top_global$gene, sep = "::")

overlap_n <- length(intersect(key_seurat, key_global))
only_seurat_n <- length(setdiff(key_seurat, key_global))
only_global_n <- length(setdiff(key_global, key_seurat))

c(overlap = overlap_n, only_seurat = only_seurat_n, only_global = only_global_n)

# also which genes dropped out of analysis
dropped <- top_seurat %>%
  mutate(key = paste(cluster, gene, sep="::")) %>%
  filter(!(key %in% key_global)) %>%
  arrange(cluster, desc(avg_log2FC))

dropped
```

```{r}
genes_to_plot <- markers %>%
  filter(!grepl("^RPL|^RPS|^MT-", gene)) %>%
  filter(p_val_adj < 0.05, pct.1 > 0.1) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), desc(pct.1 - pct.2)) %>%
  slice_head(n = 2) %>%
  pull(gene) %>%
  unique()

# use for feature plot
FeaturePlot(obj, features = genes_to_plot[1:6], reduction = "umap", order = TRUE)
VlnPlot(obj, features = genes_to_plot[1:6], ncol = 3)
```

```{r}
# Investigate cluster 3 specifically
m3 <- FindMarkers(obj, ident.1 = 3, only.pos = TRUE)
head(m3, 15)
```

```{r}
obj <- CellCycleScoring(
  obj,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes
)

VlnPlot(obj, features = c("S.Score", "G2M.Score"))
```
# let's make a sample panel to match clusters to most likely cell types 
```{r}
panel <- c("SOX2","PAX6","VIM","NES","HES1","FABP7","SLC1A3",
           "EOMES","DCX","TUBB3",
           "NEUROD6","SATB2","TBR1","SLC17A7",
           "GAD1","GAD2","DLX1","DLX2",
           "GFAP","AQP4","ALDH1L1","S100B",
           "OLIG1","OLIG2","PDGFRA","MBP","PLP1",
           "MKI67","TOP2A","CDK1")

DotPlot(obj, features = panel) + RotatedAxis()
```
# let's get a reference
```{r}
ref <- HumanPrimaryCellAtlasData()
```


# convert Seurat to SingleCellExperiment to use singleR
```{r}
sce <- as.SingleCellExperiment(obj)

pred_clust <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.main,
  clusters = Idents(obj)   # <- key fix
)

pred_clust
```

```{r}
# 1) Create mapping: cluster -> label (prefer pruned.labels if available)
cluster_to_label <- pred_clust$pruned.labels
cluster_to_label[is.na(cluster_to_label)] <- pred_clust$labels[is.na(cluster_to_label)]
names(cluster_to_label) <- rownames(pred_clust)  # "0"..."6"

# 2) Map to each cell using its cluster id
cell_clusters <- as.character(Idents(obj))        # length = #cells
cell_labels <- cluster_to_label[cell_clusters]
names(cell_labels) <- colnames(obj)              # MUST be cell names

# 3) Store in Seurat metadata
obj <- AddMetaData(obj, metadata = cell_labels, col.name = "SingleR_label")

Idents(obj) <- "seurat_clusters"   # or whatever your clustering column is
table(Idents(obj))

table(obj$SingleR_label)
table(obj$SingleR_label, obj$seurat_clusters)
```

```{r}

```


# Assign the labels back to Seurat
```{r}
FeaturePlot(obj, features = c("NEUROD6","RBFOX1","MKI67","SOX2","GFAP"), ncol = 3)
DimPlot(obj, group.by = "SingleR_label", label = TRUE)
```
# snippet
```{r}
table(obj$SingleR_label, obj$seurat_clusters)
```

# convert gene symbols to ENTREZ IDs
```{r}
sig_genes <- markers %>%
  filter(p_val_adj < 0.05) %>%
  pull(gene) %>%
  unique()

gene_df <- bitr(
  sig_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

entrez_genes <- gene_df$ENTREZID
```
# Gene ontology enrichement for biological processes
```{r}
ego <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Hs.eg.db,
  ont = "BP",            # BP / MF / CC
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)
```

```{r}

# fix spacing and wrap text
ego2 <- ego

ego2@result$Description <- str_wrap(
  ego2@result$Description,
  width = 40
)

dotplot(ego2, showCategory = 15) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.margin = margin(10, 30, 10, 10)
  )

barplot(ego2, showCategory = 20)
```

# GSEA
```{r warning=FALSE}
# --- 1) Collapse markers to 1 value per gene SYMBOL (avoid duplicates) ---
gene_list <- markers %>%
  dplyr::select(gene, avg_log2FC) %>%
  group_by(gene) %>%
  summarize(avg_log2FC = max(avg_log2FC, na.rm = TRUE), .groups = "drop") %>%  # or mean()
  arrange(desc(avg_log2FC))

# --- 2) Map SYMBOL -> ENTREZID (some will fail; that's fine) ---
gene_list_df <- bitr(
  gene_list$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# --- 3) Join + collapse to 1 value per ENTREZID (avoid 1 SYMBOL -> many ENTREZ) ---
gene_list2 <- gene_list %>%
  inner_join(gene_list_df, by = c("gene" = "SYMBOL")) %>%
  group_by(ENTREZID) %>%
  summarize(avg_log2FC = max(avg_log2FC, na.rm = TRUE), .groups = "drop") %>%  # or mean()
  arrange(desc(avg_log2FC))

# --- 4) Named ranking vector (unique names, sorted) ---
gene_vector <- gene_list2$avg_log2FC
names(gene_vector) <- gene_list2$ENTREZID
gene_vector <- sort(gene_vector, decreasing = TRUE)

stopifnot(anyDuplicated(names(gene_vector)) == 0)

# --- 5) Choose scoreType based on sign (pos if all > 0) ---
score_type <- if (all(gene_vector > 0)) "pos" else "std"

# --- RUN GSEA (GO BP) ---
gsea_res <- gseGO(
  geneList = gene_vector,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  scoreType = score_type,
  verbose = FALSE
)
```
# ridge plot of top pathways
```{r}
ridgeplot(
  gsea_res,
  showCategory = 15
) +
  theme_bw(base_size = 12)
```


