---
title: "scRNA"
author: "aidan borkan"
date: "2026-02-21"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
# installation
#install.packages("Matrix")
#BiocManager::install("GEOquery")
# ensure over 1.1.7
#install.packages("rlang", type = "binary")
#packageVersion("rlang")
reticulate::py_install("umap-learn", pip = TRUE)

# Comprehensive list
library(GEOquery)
library(dplyr)
library(Seurat)
library(patchwork)
library(Matrix)
# Annotations
library(SingleR)
library(celldex)
library(SingleCellExperiment)

gse <- getGEO("GSE75140", GSEMatrix = TRUE)
expr <- exprs(gse[[1]])
```

```{r inspect , message=TRUE, warning=FALSE}
# gse is your GEO object (from getGEO)
eset <- gse[[1]]

#expr <- exprs(eset)      # expression matrix
pheno <- pData(eset)     # sample metadata
fdata <- fData(eset)     # feature metadata (genes)
dim(expr)     # genes x samples (or probes x samples)
dim(pheno)
head(colnames(pheno))
```
# often on GEO we need the supplementary data for expr
```{r suppl, message=TRUE, warning=FALSE}
df <- read.table(
  "C:/Users/suzan/Downloads/GSE75140_hOrg.fetal.master.data.frame.txt.gz",
  header = TRUE,
  sep = "\t",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

stopifnot("cell_id" %in% colnames(df))

# Put cell IDs in rownames and drop column
rownames(df) <- df$cell_id
df$cell_id <- NULL

# DROP non-expression metadata columns (we discovered 'species' is one)
if ("species" %in% colnames(df)) df$species <- NULL

# Convert remaining columns to numeric
df[] <- lapply(df, function(x) suppressWarnings(as.numeric(x)))

# Hard check: no NAs introduced now
stopifnot(sum(is.na(as.matrix(df))) == 0)

# Confirm numeric range
range(as.matrix(df), na.rm = TRUE)
```


```{r message=FALSE, warning=FALSE}
# check rownames, I want to ensure this dataset is scRNA
# indicator would be either a barcode or cell_id
head(rownames(df))
head(colnames(df))
summary(as.vector(as.matrix(df[,1:100])))
```

```{r message=FALSE, warning=FALSE}
# Set cell IDs as rownames
rownames(df) <- df$cell_id

# Remove cell_id column
df$cell_id <- NULL

# Convert to numeric matrix
df <- as.data.frame(lapply(df, as.numeric))

# Confirm
str(df[1:5,1:5])
range(as.matrix(df), na.rm=TRUE)
```

#Step 1: read in 10x data using cellranger, in this case it's pre-processed GEO matrix
#Create an object
# transpose and create w/o re-normalizing or scaling
# store w/o zero values in memory to free RAM
```{r message=FALSE, warning=FALSE}
mat <- t(as.matrix(df))  # genes × cells

# Build object with dummy counts + data layer
zero_counts <- matrix(0, nrow=nrow(mat), ncol=ncol(mat), dimnames=dimnames(mat))
obj <- CreateSeuratObject(counts = zero_counts, project = "GSE75140")
obj <- SetAssayData(obj, assay="RNA", layer="data", new.data=mat)

# HVGs by variance (robust)
gene_var <- apply(mat, 1, var)
VariableFeatures(obj) <- names(sort(gene_var, decreasing = TRUE))[1:2000]

```

# inspect distribution to find percentiles that can guide cutoffs
```{r message=FALSE, warning=FALSE}

# Add mitochondrial % to metadata
# 1) percent.mt (optional)
# Only makes sense if gene names include MT- genes; fine to compute, but interpret cautiously
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")

# 2) Set HVGs WITHOUT loess
data_mat <- GetAssayData(obj, assay = "RNA", layer = "data")  # use data layer for your log-expression
gene_var <- apply(as.matrix(data_mat), 1, var)
VariableFeatures(obj) <- names(sort(gene_var, decreasing = TRUE))[1:2000]

# 3) Continue downstream
obj <- ScaleData(obj, features = VariableFeatures(obj))
obj <- RunPCA(obj, features = VariableFeatures(obj))
# update to umap-learn to erase warning, to use R just use uwot
obj <- RunUMAP(obj, dims = 1:20, umap.method = "umap-learn", metric = "correlation") 
obj <- FindNeighbors(obj, dims = 1:20)
obj <- FindClusters(obj)

# let's find genes detected by cell with log-exp > 0
data_mat <- GetAssayData(obj, layer = "data")

obj$genes_detected <- Matrix::colSums(data_mat > 0)
quantile(obj$genes_detected, probs = c(0.01,0.05,0.1,0.5,0.9,0.95,0.99), na.rm = TRUE)

# proxy for library size
obj$sum_logexpr <- Matrix::colSums(data_mat)
quantile(obj$sum_logexpr, probs = c(0.01,0.05,0.1,0.5,0.9,0.95,0.99), na.rm = TRUE)

# proxy for mitochondrial signal, this is GEO so this is just practice 
mt_genes <- grep("^MT-", rownames(obj[["RNA"]]), value = TRUE)
length(mt_genes)  # should be > 0; if 0, your naming isn't MT-

obj$mt_score <- Matrix::colMeans(data_mat[mt_genes, , drop = FALSE])
quantile(obj$mt_score, probs = c(0.5,0.9,0.95,0.99), na.rm = TRUE)
```


# Proxy for standard QC
```{r}
data_mat <- GetAssayData(obj, assay = "RNA", layer = "data")

obj$genes_detected <- Matrix::colSums(data_mat > 0)
obj$sum_logexpr    <- Matrix::colSums(data_mat)

mt_genes <- grep("^MT-", rownames(obj[["RNA"]]), value = TRUE)

plot_feats <- c("genes_detected", "sum_logexpr")

if (length(mt_genes) > 0) {
  obj$mt_score <- Matrix::colMeans(data_mat[mt_genes, , drop = FALSE])
  plot_feats <- c(plot_feats, "mt_score")
} else {
  message("No MT- genes found; skipping mt_score plot.")
}

VlnPlot(obj, features = plot_feats, ncol = length(plot_feats))

# note, geo data matrix does not have mt 
# higher genes_detected means higher transcriptomic complexity
```
# example of how to test if a cluster (3) is biologically meaningful: cluster and plot UMAP and see if 3 is an edge island or randomly scattered fprminmg a diffuse cloud
```{r}
FeaturePlot(obj, features = c("genes_detected"))
DimPlot(obj, label = TRUE)
```

# follow : https://satijalab.org/seurat/articles/pbmc3k_tutorial
```{r}
# problem is this is a GEO already log-normalized dataset
DefaultAssay(obj) <- "RNA"

# ------------------------------------------------------------
# 1) Mitochondrial percentage (practice only for this dataset)
# ------------------------------------------------------------
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")

# ------------------------------------------------------------
# 2) Find Highly Variable Genes (Seurat way)
# ------------------------------------------------------------
obj <- FindVariableFeatures(
  obj,
  selection.method = "dispersion",
  nfeatures = 2000
)

# Optional: visualize HVGs
top10 <- head(VariableFeatures(obj), 10)
plot1 <- VariableFeaturePlot(obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

# ------------------------------------------------------------
# 3) Scale on HVGs only
# ------------------------------------------------------------
obj <- ScaleData(
  obj,
  features = VariableFeatures(obj)
)

# ------------------------------------------------------------
# 4) PCA
# ------------------------------------------------------------
obj <- RunPCA(
  obj,
  features = VariableFeatures(obj)
)

ElbowPlot(obj, ndims = 50)

# ------------------------------------------------------------
# 5) Neighbors + Clusters
# ------------------------------------------------------------
obj <- FindNeighbors(obj, dims = 1:20)
obj <- FindClusters(obj, resolution = 0.4)

# ------------------------------------------------------------
# 6) UMAP (use uwot unless you really want Python)
# ------------------------------------------------------------
obj <- RunUMAP(
  obj,
  dims = 1:20,
  umap.method = "uwot",      # faster, no python warnings
  metric = "cosine"
)

DimPlot(obj, reduction = "umap", label = TRUE)
```

```{r}
VizDimLoadings(obj, dims = 1:2, reduction = "pca")

```
```{r}
DimPlot(obj, reduction = "pca") + NoLegend()
```
```{r}
DimHeatmap(obj, dims = 1, cells = 500, balanced = TRUE)

```
# determine dimensionality of data
```{r}
ElbowPlot(obj)
```
```{r}
obj <- RunTSNE(obj, dims = 1:20)
DimPlot(obj, reduction = "tsne", label = TRUE)
```

```{r}
#let's find markers
markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,        # keep it simple at first
  min.pct = 0.25,         # gene expressed in >=25% of cells in cluster
  logfc.threshold = 0.25  # minimum effect size
)

# Top markers per cluster
library(dplyr)
top_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10)

top_markers
```

```{r}
genes_to_plot <- c("NEUROD6", "RBFOX1", "GRIK3", "PRDM8", "NPTX1")
FeaturePlot(obj, features = genes_to_plot, reduction = "umap", ncol = 3)
FeaturePlot(obj, features = genes_to_plot, reduction = "tsne", ncol = 3)
VlnPlot(obj, features = genes_to_plot, ncol = 3)
```
```{r}
# Investigate cluster 3 specifically
m3 <- FindMarkers(obj, ident.1 = 3, only.pos = TRUE)
head(m3, 15)
```

```{r}
obj <- CellCycleScoring(
  obj,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes
)

VlnPlot(obj, features = c("S.Score", "G2M.Score"))
```
# let's make a sample panel to match clusters to most likely cell types 
```{r}
panel <- c("SOX2","PAX6","VIM","NES","HES1","FABP7","SLC1A3",
           "EOMES","DCX","TUBB3",
           "NEUROD6","SATB2","TBR1","SLC17A7",
           "GAD1","GAD2","DLX1","DLX2",
           "GFAP","AQP4","ALDH1L1","S100B",
           "OLIG1","OLIG2","PDGFRA","MBP","PLP1",
           "MKI67","TOP2A","CDK1")

DotPlot(obj, features = panel) + RotatedAxis()
```
# let's get a reference
```{r}
ref <- HumanPrimaryCellAtlasData()
```


# convert Seurat to SingleCellExperiment to use singleR
```{r}
sce <- as.SingleCellExperiment(obj)

pred_clust <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.main,
  clusters = Idents(obj)   # <- key fix
)

pred_clust
```

```{r}
# 1) Create mapping: cluster -> label (prefer pruned.labels if available)
cluster_to_label <- pred_clust$pruned.labels
cluster_to_label[is.na(cluster_to_label)] <- pred_clust$labels[is.na(cluster_to_label)]
names(cluster_to_label) <- rownames(pred_clust)  # "0"..."6"

# 2) Map to each cell using its cluster id
cell_clusters <- as.character(Idents(obj))        # length = #cells
cell_labels <- cluster_to_label[cell_clusters]
names(cell_labels) <- colnames(obj)              # MUST be cell names

# 3) Store in Seurat metadata
obj <- AddMetaData(obj, metadata = cell_labels, col.name = "SingleR_label")

Idents(obj) <- "seurat_clusters"   # or whatever your clustering column is
table(Idents(obj))

table(obj$SingleR_label)
table(obj$SingleR_label, obj$seurat_clusters)
```

```{r}

```


# Assign the labels back to Seurat
```{r}
FeaturePlot(obj, features = c("NEUROD6","RBFOX1","MKI67","SOX2","GFAP"), ncol = 3)
DimPlot(obj, group.by = "SingleR_label", label = TRUE)
```
# snippet
```{r}
table(obj$SingleR_label, obj$seurat_clusters)
```

